manuscript_v1

### 2022-10-29 manuscript.Rmd
KS: text suggestions (not comments)

Much of our knowledge on habitat selection in free-ranging animals is restricted to the breeding season. At different scales, animals assess a wide variety of aspects and features of the environment and make a series of hierarchical choices, culminating in an affinity and preference for certain habitats over others, which is termed habitat selection *(Johnson, 1980; Cody, 1985)*. Studies on habitat selection are interested in the ways in which these preferences vary across species, space, time, etc. Broad-scale selection is reflected in species’ geographic ranges and in the landscape features surrounding home ranges, while fine-scale selection is reflected in microhabitats used for foraging, reproduction, and shelter (Fretwell & Lucas, 1969; Johnson, 1980; Saab, 1999). However, the breeding season---encompassing both the actual reproductive period, as well as the period leading up to mating---dominates a large majority of these questions (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021). Moreover, most such studies target migratory species, whether it be in their breeding grounds (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021), wintering grounds (usually in the tropics; e.g., Strasser et al., 2019; Ruiz-Sanchez et al., 2017; Albert et al., 2020) or stopover sites (e.g., Papageorgiou et al., 2017; Wolfe et al., 2014; Alonso et al., 2020; Cohen et al., 2021).

This disproportionate attention is not unjustified: habitats chosen by animals can influence fitness outcomes via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use, and these have traditionally been strongly linked with reproductive success and offspring survival. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales. Therefore, investigating their selection patterns may reveal a lot about the links between organisms and habitat and landscape features. Nevertheless, habitat dynamics during transitional extra-breeding periods can be of great importance as well, given their associated high stress and consequence as an important link in the chain. In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in these periods. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a). Things are not made easier by the arrival of migrants from the north, resulting in a higher overall resource consumption. Despite slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014), the post-breeding period remains understudied, at least in part due to the much lower vocal activity of birds in this period.

Species vary in their degree of selectivity from generalists to specialists, with the latter being less flexible and more susceptible to change (Devictor et al., 2008; Owens and Bennett, 2000). Many migrating land birds generally considered to be insectivorous, at least during the breeding period, cope with the food pressure in the post-breeding period by exhibiting a high level of foraging and dietary flexibility (Herrera, 1978; Gauthreaux Jr, 1982; Parrish, 1997, 2000; McWilliams and Karasov, 2001; Salewski and Jones, 2006). These birds may become facultatively frugivorous in autumn, allowing them to rapidly build up the fat reserves vital for migration, as well as making them less reliant on the declining arthropods (Bairlein and Simons, 1995; Faaborg, 2002; Bairlein, 2003; Ottich and Dierschke, 2003; Newton, 2010; Hernández, 2009). However, an important distinction between temperate forests of the Old and New Worlds is that there are no specialised obligate frugivorous species in the Old World temperate avifauna, and more importantly, European temperate forests have a much lower diversity of fruiting plants than American ones. Among the most widespread fruiting species are bramble *Rubus* spp., bilberry and blueberry *Vaccinium* spp., elderberry *Sambucus nigra*, rowan *Sorbus aucuparia* and alder buckthorn *Frangula alnus*. These species are typical of hedgerows, wood edges and disturbed patches, and old growth mixed forests, but are scarcely found in the large, compact and structurally poor stands of present-day Central European forests. Therefore, the dietary options for Central European forest birds, particularly residents, are rather limited. Selection patterns are likely to be more dependent on structural characteristics of the habitat, such as heterogeneity and diversity, and whether the species chooses to adapt to varying features or to master one.

Oak forests are widely believed to be the most biodiversity-rich, supporting a highly diverse arthropod community which in turn supports a larger vertebrate community. These forests also have more heterogeneous canopies and understoreys than the structurally uniform coniferous forests. As a result, birds during the post-breeding period could show a strong foraging preference for oak forest patches. Terrestrial arthropods have been found to comprise a surprisingly large proportion of the diet of some wintering birds even in winter. For instance, Gibb (1960) found that birds in winter consumed around 50% of the stock of the eucosmid *Ernarmonia conicolana* which is known to be an important food source of blue tits. In spruce forests, tits (Paridae) commonly consume Lepidoptera, Hemiptera, Coleoptera and spiders (Betts, 1955). Other observations from forests suggest that spiders and insects including eggs, larvae, pupae and adults in diapause are also important winter food sources (Williams and Batzli, 1979; Jansson and Brömssen, 1981; Laurent, 1986). There is strong evidence that birds can indirectly impact terrestrial plant communities through arthropod predation (Van Bael et al., 2008; Mooney et al., 2010; Mäntylä et al., 2011), although few such studies in temperate regions have explored these effects in winter. Strictly insectivorous birds may be able to monopolise the arthropod food sources, while omnivores might enjoy the advantages that come with being generalists.

@@

These various strategies and interactions are usually linked with habitat characteristics such as canopy cover, heterogeneity, tree density and species richness (Vitz and Rodewald, 2007). Such characteristics are commonly compared with bird abundances or densities to analyse habitat selection. Berg (1997) found that bird abundances correlated positively with the presence of deciduous trees and with tree diameter, while among conifers the Norway spruce *Picea abies* was preferred over the pine *Pinus sylvestris*. Spatial segregation of habitats at a relatively fine-grained scale is suggested to allow for the co-occurrence of more species. However, in addition to hosting a low diversity of fruiting plant species, Central European forests are also less spatially heterogenous. Most modern-day forests in Central Europe have been restructured towards commercially exploited plantations dominated by spruce that form compact growths with a closed above-tree layer at maturity (Fanta, 2007). On the other hand, intense management and disturbance regimes bring in heterogeneity, but do not facilitate biodiversity for obvious reasons. In these circumstances, studies on habitat selection can provide crucial information that can better advise conservation action in the ever-globalising world.

In this study, we investigated patterns of habitat selection exhibited by temperate forest bird communities in the post-breeding period at a fine-grained temporal (weeks) and spatial (patches) resolution. The main objectives were to identify the habitat features that are tracked most prominently in this period of high stress (i.e., the autumn-winter transition), to assess the patterns and magnitude of habitat selection as well as their possible temporal dynamism with the seasonal transition, and to explore the possibilities of ecological interactions and strategies.

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated and commercially exploited patch of Central European temperate forest (*Branisovsky les*) in České Budějovice, Czech Republic. The forest is located on the western edge of the city (48°58'55.672"N, 14°25'22.907"E), and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is very patchy owing to the long history of forestry regimes as well as its connection with rural and military Czech history. The patches range from pure spruce plantations to mixed stands dominated by oak. As such, the tree community consists mostly of conifers like Norway spruce *Picea abies*, Scots pine *Pinus sylvestris*, European larch *Larix decidua* and silver fir *Abies alba*, and broad-leaved species such as pedunculate/European oak *Quercus robur*, sessile/Cornish oak *Q. petraea*, northern red oak *Q. rubra*, silver birch *Betula pendula*, quaking/Eurasian aspen *Populus tremula*, common/European beech *Fagus sylvatica*, rowan or European mountain ash *Sorbus aucuparia*, small-leaved lime or littleleaf linden *Tilia cordata*, common/European black alder *Alnus glutinosa*, Norway maple *Acer platanoides* and sycamore maple *A. pseudoplatanus*. The commonly found shrub species are alder buckthorn *Frangula alnus*, rowan *Sorbus aucuparia* and black/European elderberry *Sambucus nigra*, while ground layer vegetation is restricted to graminoids like sedges *Carex* spp. (mostly *C. brizoides*), bramble *Rubus* sp. and European blueberry *Vaccinium myrtillus*.

The entire forest stretch---with the exception of some parts either too disturbed or having restricted access---was effectively covered by selecting 32 non-random, evenly distributed circular sampling points of 30 m radius (Figure 1). The total forest area under consideration was thus around 400 ha, while the absolute cumulative area covered by the sampling points was approximately 9 ha, as they were chosen such that each point was at least 250 m away from its neighbours (although this was lowered to 200 m in three cases). Clear cuts were avoided as much as possible. We ensured that the selected set of points represented sufficiently well the natural range and diversity of habitat characteristics in the forest.

@@

Vegetation features thought to be relevant to birds in the post-breeding period were selected and measured to initially characterise the various habitat patches. The habitat class of each sampling point was recorded as ‘Interior’, ‘Road’ or ‘Edge’ according to whether or not asphalt roads and forest-to-farmland transitions occurred within 150 m from the centre of the point. This was done using an aerial map (Mapy.cz) in which asphalt roads, farmlands and forests are all clearly visible and distinguishable. At each of the 32 points, the tree, shrub and herb species were identified, and the vertical strata to which they belonged were noted. The species information was used to calculate plant species richness and diversity (Shannon’s diversity index; Shannon, 1948). Regardless of taxonomic and functional identity, plants of height $\le$ 1 m above the ground classified into the ‘Herb’ layer (this therefore included both woody- [like *Rubus*] and non-woody stemmed [like graminoids] plants, but did not exclude graminoid leaves that extended beyond 1 m from the ground); plants 1--4 m in height belonged to the ‘Understorey’ (this included shrubs like *S. aucuparia* and *F. alnus* as well as young individuals of trees like *P. abies*); and plants $\ge$ 4 m in height belonged to the ‘Canopy’ (this occasionally included *S. aucuparia*). Counts of ‘Canopy’ and ‘Understorey’ plants were taken and used to calculate density per 100 m^2. Counts were not taken for the ‘Herb’ layer due to the difficulty in some cases in determining what constitutes a single stem and since it is unlikely that birds track ground layer vegetation at the level of individual stems. Instead, the species that covered $\gt$ 50% of the ground was considered to be the ‘Dominant Ground Vegetation’ (in some cases, this was moss) and this variable was also used in analyses. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch (Berdugo et al., 2018; Cornelissen et al., 2007).

We also calculated the proportion of deciduous tress in the 'Canopy' as this has been shown to be a more important factor than measures of tree diversity (Berg, 1997). Canopy cover at a sampling point was measured using canopy photographs taken at three equidistant sub-points within 10--15 m of the centre, which were then analysed using ImageJ software (Rasband, 1997) by converting to binary black-and-white images and calculating the percentage of cover across the whole image (see Forester, 2011; Figure S1). The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). These measurements were taken once in each sampling week. Since most deciduous trees bore leaves even in the last week of sampling, the 13 weekly measurements of both mean canopy cover and canopy heterogeneity for each sampling point were averaged to provide more accurate measures of both. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period.

@@

We surveyed bird 'abundance' by means of fixed-radius point counts, using detections as a proxy for abundance. While detections are not an accurate measure of true abundance, with many vegetation characteristics and behavioural traits affecting detectability of birds, it was sufficient for our purposes since we were not interested in estimating true population size, and the sampling points did not vary highly in terms of vegetation characteristics hindering detectability or detectability of different bird communities (rather, most birds have similarly low levels of vocal activity in the post-breeding period). Point counts were done at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020), thus covering the post-breeding period from late summer till early autumn at a fine-grained temporal resolution. Each point being located at least 250 m from its neighbours assured spatiotemporal independence of consecutive point counts. Sampling was done on 2--3 days in the middle of the calendar week while the remaining days served to delineate and separate two successive ‘temporal units’. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point (Table 1).

We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen during a period of 20 minutes, divided into four intervals of five minutes each. Additionally, complementary information for each detection, like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and $\gt$ 50 m. Additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the sampling point) at any time during the count, signifying the bird’s ‘use’ of the habitat. All birds were identified to species level when possible, but some closely related cryptics such as *Phylloscopus* spp., *Certhia* spp., *Regulus* spp. and *Dendrocopos* spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as ‘spuhs’.

Prior to analysis, bird species were classified based on feeding guild as well as migration strategy. Feeding guild was determined using data from an extensive network of European ornithologists (Gregory et al., 2007, pt. Appendix 1) complemented with species descriptions from Birds of the World (Billerman et al., 2020). Migration strategy was assigned following Czech ornithological data; species of which $\gt$ 75% of the Czech breeding population migrates either short or long distances (categories 3 and 4 in Cepák et al., 2008) were considered as complete migrants for the purposes of this study; those with ‘regular and abundant wintertime occurrence in the Czech Republic’ were considered partial migrants (category 2 in Cepák et al., 2008; see also Št’astnỳ and Hudec, 2011); and the residents of category 1 (Cepák et al., 2008) were considered true residents.

@@

To assess patterns of habitat selection by birds, we excluded birds not identified to genus level, those detected flying over the forest and those that did not enter the 30 m radius in point counts. The measured habitat variables without the ones excluded due to correlations represented the pool of candidate predictors. This set of plausible predictors was then used to build generalised linear mixed models (GLMM) following an information-theoretic approach (Burnham and Anderson, 1998). That is, the Akaike Information Criterion (Akaike, 1973, 1974), and more specifically the corrected AIC (AIC~C~; Hurvich and Tsai, 1989), was used to judge model fit.

Analyses were conducted at two main levels of response. At the highest level, we looked for patterns of habitat selection in the whole bird community, by summing all the individual species detections per point count to obtain one 'bird abundance' value for each count, which was then used as the response variable. GLMMs were constructed using the *lme4* package (Bates et al., 2015), starting from a null model containing observer as a fixed effect and sampling points and days as crossed random effects, then building up. Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day in order to account for the random variability in each unique day. Residuals were modelled as coming from a Poisson distribution with a log link function. Next, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in) were added. Following this, finally the focal variables were added. For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package (Bolker et al., 2020), with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with ‘Week’ was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Model diagnostics of all models were explored using the *DHARMa* package (Hartig, 2020) for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected.

@@

A total of 7,367 detections of 69 bird species were made, of which 62% (4527) did not meet the criteria for inclusion in the analysis. Of the 2,840 detections of 45 species anlaysed, 78% (2,209) were purely auditory, 7% (201) were purely visual and 15% (430) were both, and this did not change majorly with weeks or across points (Figure S3). The average number of detections per point count varied from 5 in Week 1 to 9 in Week 13 (Figure 2). Among 494 records of birds foraging, just two were of frugivory.

@@

There was an overall increase in bird abundance towards the end of the study period in all sampling points---across the five categories of ground vegetation---but the magnitude of increase was starkly different in each. Points with *Rubus* sp. and those with *Vaccinium* sp. as the dominant ground vegetation had intermediate values of both average abundance and increase in abundance with time. Points with graminoids in the ground layer had the lowest overall abundance as well as the lowest increase (Figure 3). Points with ground layers comprising moss and bare ground (category ‘Bare’) both had low abundances in the beginning but towards the end had the highest abundances, i.e., they showed the strongest increases. Bird abundance related positively and negatively with log-transformed variables of canopy heterogeneity and tree density respectively (Figure 3) and these trends did not change noticeably over the course of the study period.

>>>>

Much of our knowledge on habitat selection of free-living animals is restricted to the breeding season. At different scales, animals assess a wide variety of aspects and features of the environment and make a series of hierarchical choices, culminating in an affinity and preference for certain habitats over others, which is termed habitat selection *(Johnson, 1980; Cody, 1985)*. Studies on habitat selection are interested in the ways in which these preferences vary across species, space, time, etc. Broad-scale selection is reflected in species' geographic ranges and in the landscape features surrounding home ranges, while fine-scale selection is reflected in microhabitats used for foraging, reproduction, and shelter (Fretwell & Lucas, 1969; Johnson, 1980; Saab, 1999). However, the breeding season---encompassing both the actual reproductive period, as well as the period leading up to mating---dominates a large majority of these questions (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021). Moreover, most such studies target migratory species, whether it be in their breeding grounds (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021), wintering grounds (usually in the tropics; e.g., Strasser et al., 2019; Ruiz-Sanchez et al., 2017; Albert et al., 2020) or stopover sites (e.g., Papageorgiou et al., 2017; Wolfe et al., 2014; Alonso et al., 2020; Cohen et al., 2021).

This disproportionate attention is not unjustified: habitats chosen by animals can influence fitness outcomes via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use, and these have traditionally been strongly linked with reproductive success and offspring survival. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales. Nevertheless, habitat dynamics during transitional extra-breeding periods can be of great importance as well, given their associated high stress and consequence as an important link in the chain. In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in these periods. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a). Things are not made easier by the arrival of migrants from the north, resulting in a higher overall resource consumption. Despite slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014), the post-breeding period remains understudied, at least in part due to the much lower vocal activity of birds in this period.

Species vary in their degree of selectivity from generalists to specialists, with the latter being less flexible and more susceptible to change (Devictor et al., 2008; Owens and Bennett, 2000). Many migrating land birds which are generally considered to be insectivorous, at least during the breeding period, face food shortage in the post-breeding period (Herrera, 1978; Gauthreaux Jr, 1982; Parrish, 1997, 2000; McWilliams and Karasov, 2001; Salewski and Jones, 2006). These birds may become facultatively frugivorous and more flexible in foraging in autumn, allowing them to rapidly build up the fat reserves vital for migration, as well as making them less reliant on the declining arthropods (Bairlein and Simons, 1995; Faaborg, 2002; Bairlein, 2003; Ottich and Dierschke, 2003; Newton, 2010; Hernández, 2009). However, an important distinction between temperate forests of the Old and New Worlds is that there are no specialised obligate frugivorous species in the Old World temperate avifauna, and more importantly, European temperate forests have a much lower diversity of fruiting plants than American ones. Among the most widespread fruiting species are bramble *Rubus* spp., bilberry and blueberry *Vaccinium* spp., elderberry *Sambucus nigra*, rowan *Sorbus aucuparia* and alder buckthorn *Frangula alnus*. These species are typical of hedgerows, wood edges and disturbed patches, and old growth mixed forests, but are scarcely found in the large, compact and structurally poor stands of present-day Central European forests. Therefore, the dietary options for Central European forest birds, particularly residents, are rather limited. Selection patterns are likely to be more dependent on structural characteristics of the habitat, such as heterogeneity and diversity, and whether the species chooses to adapt to varying features or to master one.

Temperate oak forests are widely believed to be the most biodiversity-rich, supporting a highly diverse arthropod community which in turn supports a larger vertebrate community. These forests also have more heterogeneous canopies and understoreys than the structurally uniform coniferous forests. As a result, birds during the post-breeding period could show a strong foraging preference for oak forest patches. Terrestrial arthropods have been found to comprise a surprisingly large proportion of the diet of some wintering birds even in winter. For instance, Gibb (1960) found that birds in winter consumed around 50% of the stock of the eucosmid *Ernarmonia conicolana* which is known to be an important food source of blue tits. In spruce forests, tits (Paridae) commonly consume Lepidoptera, Hemiptera, Coleoptera and spiders (Betts, 1955). Other observations from forests suggest that spiders and insects including eggs, larvae, pupae and adults in diapause are also important winter food sources (Williams and Batzli, 1979; Jansson and Brömssen, 1981; Laurent, 1986). There is strong evidence that birds can indirectly impact terrestrial plant communities through arthropod predation (Van Bael et al., 2008; Mooney et al., 2010; Mäntylä et al., 2011), although few such studies in temperate regions have explored these effects in winter. Strictly insectivorous birds may be able to monopolise the arthropod food sources, while omnivores might enjoy the advantages that come with being generalists.

@@

These various strategies and interactions are usually linked with habitat characteristics such as canopy cover, heterogeneity, tree density and species richness (Vitz and Rodewald, 2007). Bird abundances or densities are compared with these characteristics to analyse habitat selection. Berg (1997) found that bird abundances correlated positively with the presence of deciduous trees and with tree diameter, while among conifers the Norway spruce *Picea abies* was preferred over the pine *Pinus sylvestris*. Spatial segregation of habitats at a relatively fine-grained scale is suggested to allow for the co-occurrence of more species. However, in addition to hosting a low diversity of fruiting plant species, Central European forests are also less spatially heterogenous. Most modern-day forests in Central Europe have been restructured towards commercially exploited plantations dominated by spruce that form compact growths with a closed above-tree layer at maturity (Fanta, 2007). On the other hand, intense management and disturbance regimes bring in heterogeneity, but do not facilitate biodiversity for obvious reasons. In these circumstances, studies on habitat selection can provide crucial information that can better advise conservation action in the ever-globalising world.

In this study, we investigated patterns of habitat selection exhibited by forest bird communities in the post-breeding period at a fine-grained temporal (weeks) and spatial (patches) resolution in Central Europe. The main objectives were to identify the habitat features that are tracked most prominently in this period of high stress (the autumn-winter transition), to assess the patterns and magnitude of habitat selection as well as their possible temporal dynamism with the seasonal transition, and to explore the possibilities of ecological interactions and strategies.

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated and commercially used patch of Central European temperate mixed forest (*Branisovsky les*) close to České Budějovice, Czech Republic. The forest is located on the western edge of the city (48°58'55.672"N, 14°25'22.907"E), and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is very patchy owing to the long history of forestry regimes as well as its connection with rural and military Czech history. The patches range from pure Norway spruce *Picea abies* plantations to mixed stands dominated by oaks (*Quercus* spp.). As such, the tree community consists mostly of conifers like Norway spruce, Scots pine *Pinus sylvestris*, European larch *Larix decidua* and silver fir *Abies alba*, and broad-leaved species such as pedunculate/European oak *Quercus robur*, sessile/Cornish oak *Q. petraea*, northern red oak *Q. rubra*, silver birch *Betula pendula*, quaking/Eurasian aspen *Populus tremula*, common/European beech *Fagus sylvatica*, rowan or European mountain ash *Sorbus aucuparia*, small-leaved lime or littleleaf linden *Tilia cordata*, common/European black alder *Alnus glutinosa*, Norway maple *Acer platanoides* and sycamore maple *A. pseudoplatanus*. The commonly found shrub species are alder buckthorn *Frangula alnus*, rowan *Sorbus aucuparia* and black/European elderberry *Sambucus nigra*, while ground layer vegetation is restricted to graminoids like sedges *Carex* spp. (mostly *C. brizoides*), bramble *Rubus* sp. and European blueberry *Vaccinium myrtillus*.

The entire forest stretch---with the exception of some parts either too disturbed or having restricted access---was effectively covered by selecting 32 non-random, evenly distributed circular sampling points of 30 m radius (Figure 1). They were chosen such that each point was at least 250 m away from its neighbours, so as to ensure spatiotemporal independence of consecutive point counts. Thus, the absolute cumulative area covered by the sampling points was approximately 9 ha. Clear cuts were avoided as much as possible. The selected set of points represented well the natural range and diversity of habitat characteristics in the entire forest.

@@

Vegetation features potentially relevant to birds in the post-breeding period were selected and measured to initially characterise the various habitat patches. The habitat class of each sampling point was recorded as 'Interior', 'Road' or 'Edge' according to whether or not asphalt roads and forest-to-farmland transitions occurred within 150 m from the centre of the point. This was done using an aerial map (Mapy.cz). At each of the 32 points, the tree, shrub and herb species were identified, and the vertical strata to which they belonged were noted. The species information was used to calculate plant species richness and diversity (Shannon's diversity index; Shannon, 1948). Regardless of taxonomic and functional identity, plants of height $\le$ 1 m above the ground classified into the 'Herb' layer (this therefore included both woody- [like *Rubus*] and non-woody stemmed [like graminoids] plants, but did not exclude graminoid leaves that extended beyond 1 m from the ground); plants 1--4 m in height belonged to the 'Understorey' (this included shrubs like *S. aucuparia* and *F. alnus* as well as young individuals of trees like *P. abies*); and plants $\ge$ 4 m in height belonged to the 'Canopy' (this occasionally included *S. aucuparia*). Counts of 'Canopy' and 'Understorey' plants were taken and used to calculate density per 100 m\^2. Counts were not taken for the 'Herb' layer due to the difficulty in some cases in determining what constitutes a single stem and since it is unlikely that birds track ground layer vegetation at the level of individual stems. Instead, the species that covered $\gt$ 50% of the ground was considered to be the 'Dominant Ground Vegetation' (in some cases, this was moss) and this variable was also used in analyses. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch (Berdugo et al., 2018; Cornelissen et al., 2007).

We also calculated the proportion of deciduous trees in the 'Canopy' as this has been shown to be a more important factor than measures of tree diversity (Berg, 1997). Canopy cover at a sampling point was measured using canopy photographs taken at three equidistant sub-points within 10--15 m of the centre, which were then analysed using ImageJ software (Rasband, 1997) by converting to binary black-and-white images and calculating the percentage of cover across the whole image (see Forester, 2011; Figure S1). The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). These measurements were taken once in each sampling week. Since most deciduous trees bore leaves even in the last week of sampling, the 13 weekly measurements of both mean canopy cover and canopy heterogeneity for each sampling point were averaged to provide more accurate measures of both. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period.

@@

We surveyed bird 'abundance' by means of fixed-radius point counts, using detections as a proxy for abundance. While detections are not an accurate measure of true abundance, with many vegetation characteristics and behavioural traits affecting detectability of birds, it was sufficient for our purposes since we were not interested in estimating true population size, and the sampling points did not vary highly in terms of vegetation characteristics hindering detectability or detectability of different bird communities (rather, most birds have similarly low levels of vocal activity in the post-breeding period). Point counts were done at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020), thus covering the post-breeding period from late summer till early autumn at a fine-grained temporal resolution. Sampling was done on 2--3 days in the middle of the calendar week while the remaining days served to delineate and separate two successive 'temporal units'. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point (Table 1).

We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen during a period of 20 minutes, divided into four intervals of five minutes each. Additionally, complementary information for each detection, like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and $\gt$ 50 m. Additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the sampling point) at any time during the count, signifying the bird's 'use' of the habitat. All birds were identified to species level when possible, but some closely related cryptics such as *Phylloscopus* spp., *Certhia* spp., *Regulus* spp. and *Dendrocopos* spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as 'spuhs'.

Prior to analysis, bird species were classified based on feeding guild as well as migration strategy. Feeding guild was determined usingGregory et al., 2007 (pt. Appendix 1) and Birds of the World (Billerman et al., 2020). Migration strategy was assigned following Czech ornithological data; species of which $\gt$ 75% of the Czech breeding population migrates either short or long distances (categories 3 and 4 in Cepák et al., 2008) were considered as complete migrants for the purposes of this study; those with 'regular and abundant wintertime occurrence in the Czech Republic' were considered partial migrants (category 2 in Cepák et al., 2008; see also Št'astnỳ and Hudec, 2011); and the residents of category 1 (Cepák et al., 2008) were considered true residents.

@@

To assess patterns of habitat selection by birds, we first excluded birds not identified to genus level, those detected flying over the forest and those that did not enter the 30 m radius in point counts. The measured habitat variables without the ones excluded due to correlations represented the pool of candidate predictors. This set of plausible predictors was then used to build generalised linear mixed models (GLMMs) following an information-theoretic approach (Burnham and Anderson, 1998). That is, the Akaike Information Criterion (AIC; Akaike, 1973, 1974), and more specifically the corrected AIC or AIC~C~ (Hurvich and Tsai, 1989), was used to judge model fit.

Analyses were conducted at two main levels of response. At the highest level, we looked for patterns of habitat selection in the whole bird community, by summing all the individual species detections per point count to obtain one 'bird abundance' value for each count, which was then used as the response variable. GLMMs were constructed using the *lme4* package (Bates et al., 2015), starting from a null model containing observer as a fixed effect and sampling points and days as crossed random effects, then building up. Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day in order to account for the random variability in each unique day. Residuals were modelled as coming from a Poisson distribution with a log link function. Next, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in) were added. Following this, finally the focal variables were added. For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package (Bolker et al., 2020), with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Model diagnostics of all models were explored using the *DHARMa* package (Hartig, 2020) for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected.

@@

A total of 7,367 detections of 69 bird species were made, of which 62% (4,527) did not meet the criteria for inclusion in the analysis. Of the 2,840 detections of 45 species anlaysed, 78% (2,209) were purely auditory, 7% (201) were purely visual and 15% (430) were both, and this did not change majorly with weeks or across points (Figure S3). The average number of detections per point count varied from 5 in Week 1 to 9 in Week 13 (Figure 2). Among 494 records of birds foraging, just two were of frugivory.

@@

There was an overall increase in bird abundance towards the end of the study period in all sampling points---across the five categories of ground vegetation---but the magnitude of increase was starkly different in each. Points with *Rubus* sp. and those with *Vaccinium* sp. as the dominant ground vegetation had intermediate values of both average abundance and increase in abundance with time. Points with graminoids in the ground layer had the lowest overall abundance as well as the lowest increase (Figure 3). Points with ground layers comprising moss and bare ground (category 'Bare') both had low abundances in the beginning but towards the end had the highest abundances, i.e., they showed the strongest increases. Bird abundance related positively and negatively with log-transformed variables of canopy heterogeneity and tree density respectively (Figure 3) and these trends did not change noticeably over the course of the study period.
